Transferring Value Functions by Variational Methods

### 2018-05-17 NIPS 2018/main.tex
Minor changes to Section 5

We consider an agent navigating in the rooms environment depicted in Fig. \ref{}. The agent starts in the bottom-left corner and must move from one room to another to reach the goal position in the top-right corner. The rooms are connected by small doors whose positions are unknown to the agent. The state-space is modeled as a $10 \times 10$ continuous grid, while the action-space is the set of $4$ movement directions (up, right, down, left). After each action, the agent moves by $1$ in the chosen direction and the final position is corrupted by Gaussian noise with $0.2$ standard deviation. In case the agent hits a wall, its position remains unchanged. The reward is $1$ when reaching the goal (after which the process terminates) and $0$ otherwise, while the discount factor is $\gamma = 0.99$. In this experiment, we consider linearly parameterized $Q$-functions with $121$ equally-spaced radial basis features.

Since the agent does not know where the doors are located in advance and receives only very sparse feedback, it must efficiently explore the environment to figure out (i) their positions, and (ii) how to reach the goal. While this might be a complicated problem for plain RL, our transfer algorithm should be able to quickly figure out the door positions. In fact, notice that, although different, the optimal $Q$-functions for all tasks share some similarities. For example, once the agent has passed the last door before the goal, the $Q$-values are exactly the same in all tasks. This does not hold for positions nearby the start state. However, it is clear that there should be a preference over actions up and right, rather than down and left (which are worse in all tasks).\todo{Good candidate to be removed}

@@

\draw [thick] (0,0) rectangle (5,5);
\draw [line width=0.5mm] (1.25,5/3) -- (5,5/3);
\draw [line width=0.5mm] (0,5/3) -- (0.75,5/3);
\draw [line width=0.5mm] (0,10/3) -- (3.25,10/3);
\draw [line width=0.5mm] (3.75,10/3) -- (5,10/3);
\end{scope}
\end{tikzpicture} 
    \caption{}
  \end{subfigure}
  \begin{subfigure}[b]{0.27\textwidth}
    \includegraphics[trim=0.5cm 0cm 1.8cm 1.3cm,clip=true,height=2.9cm]{images/three-room/lrev.pdf}
    \caption{}
  \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
    \includegraphics[trim=0.5cm 0cm 1.6cm 1.3cm,clip=true,height=2.9cm]{images/three-room-gen/lrev.pdf}
    \caption{}
  \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
    \includegraphics[trim=0.5cm 0cm 1.6cm 1.3cm,clip=true,height=2.9cm]{images/likelihood/lrew.pdf}
    \caption{}
  \end{subfigure}
  \caption{(a) the rooms environment, (b) transfer from $10$ source tasks with both doors moving, (c) transfer from $10$ source tasks with only one door moving, and (d) transfer performance as a function of how likely the target task is according to the prior.}
  \label{fig:dipfdbsgraphplots}
\end{figure*}

In order to prove that our guesses are correct, we generate a set of $50$ source tasks for the three-room environment of Fig. \ref{} by sampling both door positions uniformly in the allowed space, and solve all of them by directly minimizing the TD error as presented in Sec. \ref{sec:td}. Then, we use our algorithms to transfer from $10$ source tasks sampled from the previously generated set. The average return over the last $50$ learning episodes as a function of the number of iterations is shown in Fig. \ref{}. Each curve is the result of $20$ independent runs, each resampling the target and source tasks, with $95\%$ confidence intervals. Further details on the parameters adopted in this experiment are given in App. \ref{}. As expected, the no-transfer (NT) algorithm fails at learning the task in so few iterations due to the limited exploration provided by an $\epsilon$-greedy policy. On the other hand, all our algorithms achieve a significant speed-up and are able to converge to the optimal performance in few iterations, with GVT being slightly slower. Interestingly, we notice that there is no advantage in adopting more than $1$ component for the posterior in MGVT. This is intuitive since, when the algorithm quickly figures out which task is being solved, it will move all components in the same direction.

To better understand the differences between GVT and MGVT, we now consider transferring from a slightly different distribution than the one from which target tasks are drawn. We generate again $50$ source tasks but this time with the bottom door fixed at the center and the other one moving. Then, we repeat the previous experiment, allowing both doors to move when sampling target tasks. The results are shown in Fig. \ref{}. Interestingly, MGVT seems almost unaffected by this change, proving that is has sufficient representation power to generalize to slightly different task distributions. The same does not hold for GVT, which now is not able to solve many of the sampled target tasks, as can be noticed from the higher variance. This proves again that assuming Gaussian distributions can pose severe limitations in our transfer settings.

Finally, we analyze the transfer performance as a function of how likely the target task is according to the prior. We consider a two-room version of the environment of Fig. \ref{}. Differently from before, we generate tasks by sampling the door position from a Gaussian with mean $5$ and standard deviation $1.8$, so that tasks where the door is near the sides are very unlikely. Fig. \ref{} shows the performance reached by GVT and 1-MGVT at fixed iterations as a function of how likely the target task is according to such distribution. As expected GVT achieves poor performance on very unlikely tasks, even after many iterations. In fact, estimating a single Gaussian distribution definitely entails some information loss, especially about the unlikely tasks. On the other hand, MGVT keeps such information and, consequently, performs much better. Perhaps not surprisingly, MGVT reaches the optimal performance in $4k$ iterations no matter what task is being solved.

@@

We now consider two well-known classic control environments: Cartpole and Mountain Car \citep{sutton1998reinforcement}. For both, we generate $20$ source tasks by uniformly sampling their physical parameters (cart mass, pole mass, pole length for Cartpole and car speed for Mountain Car). We parameterize $Q$-functions using neural networks with $1$-layer of $32$ hidden units for Cartpole and $64$ for Mountain Car. A better description of these two environments and their parameters is given in App. \ref{}. In this experiment, we use a Double Deep Q-Network \cite{van2016deep} to provide a stronger no-transfer baseline for comparison. The results (same settings of Sec. \ref{sec:gw}) are shown in Fig. \ref{} and \ref{}. For Cartpole (Fig. \ref{}), all transfer algorithms are almost zero-shot. This is expected since, although we vary the system parameters in a wide range, the optimal $Q$-values of states near the balanced position are almost the same for all tasks. On the contrary, in Mountain Car (Fig. \ref{}) the optimal $Q$-functions become really different when changing the car speed. This causes GVT to struggle to solve the target task, while 1-MGVT achieves a good jump-start and convergences in less iterations.

@@

\begin{figure*}[t]
  \begin{subfigure}[b]{0.15\textwidth}
    \includegraphics[trim=0.0cm 0.0cm 0.0cm 0.0cm,clip=true,height=2.6cm, width=2.0cm]{images/maze/maze6.png}
    \caption{}
  \end{subfigure}
  \begin{subfigure}[b]{0.27\textwidth}
    \includegraphics[trim=0.5cm 0cm 1.8cm 1.3cm,clip=true,height=2.9cm]{images/cartpole/lrev.pdf}
    \caption{}
  \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
    \includegraphics[trim=0.5cm 0cm 1.6cm 1.3cm,clip=true,height=2.9cm]{images/mountaincar/lrev.pdf}
    \caption{}
  \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
    \includegraphics[trim=0.5cm 0cm 1.6cm 1.3cm,clip=true,height=2.9cm]{images/maze/lrev.pdf}
    \caption{}
  \end{subfigure}
  \caption{(a) Sample Maze (Maze 6), (b) transfer from $10$ sources in Cartpole, (c) transfer from $10$ source tasks in Mountain Car, and (d) transfer from $10$ sources in Mazes 6 and 13---see App. \ref{}.}
  \label{fig:ccmazeplots}
\end{figure*}

>>>>

We consider an agent navigating in the environment depicted in Figure~\ref{f:rooms}. The agent starts in the bottom-left corner and must move from one room to another to reach the goal position in the top-right corner. The rooms are connected by small doors whose locations are unknown to the agent. The state-space is modeled as a $10 \times 10$ continuous grid, while the action-space is the set of $4$ movement directions (up, right, down, left). After each action, the agent moves by $1$ in the chosen direction and the final position is corrupted by Gaussian noise $\mathcal{N}(0,0.2)$. In case the agent hits a wall, its position remains unchanged. The reward is $1$ when reaching the goal (after which the process terminates) and $0$ otherwise, while the discount factor is $\gamma = 0.99$. In this experiment, we consider linearly parameterized $Q$-functions with $121$ equally-spaced radial basis features.

Since the agent does not know the locations of the doors in advance and receives only very sparse feedback, it must efficiently explore the environment to figure out (i) their positions, and (ii) how to reach the goal. While this might be a complicated problem for plain RL, our transfer algorithm should be able to figure out the door positions quickly. In fact, notice that, although different, the optimal $Q$-functions for all tasks share some similarities. For example, once the agent has passed the last door before the goal, the $Q$-values are exactly the same in all tasks. The same does not hold for positions nearby the starting state. However, it is clear that there should be a preference over actions \textit{up} and \textit{right}, rather than \textit{down} and \textit{left} (which are worse in all tasks).\todo{Good candidate to be removed}

@@

\draw [thick] (0,0) rectangle (5,5);
\draw [line width=0.5mm] (1.25,5/3) -- (5,5/3);
\draw [line width=0.5mm] (0,5/3) -- (0.75,5/3);
\draw [line width=0.5mm] (0,10/3) -- (3.25,10/3);
\draw [line width=0.5mm] (3.75,10/3) -- (5,10/3);
\end{scope}
\end{tikzpicture} 
    \caption{}\label{f:rooms}
  \end{subfigure}
  \begin{subfigure}[b]{0.27\textwidth}
    \includegraphics[trim=0.5cm 0cm 1.8cm 1.3cm,clip=true,height=2.9cm]{images/three-room/lrev.pdf}
    \caption{}\label{f:two_doors}
  \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
    \includegraphics[trim=0.5cm 0cm 1.6cm 1.3cm,clip=true,height=2.9cm]{images/three-room-gen/lrev.pdf}
    \caption{}\label{f:one_door}
  \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
    \includegraphics[trim=0.5cm 0cm 1.6cm 1.3cm,clip=true,height=2.9cm]{images/likelihood/lrew.pdf}
    \caption{}\label{f:likelihood}
  \end{subfigure}
  \caption{(a) the rooms environment, (b) transfer from $10$ source tasks with both doors moving, (c) transfer from $10$ source tasks with only one door moving, and (d) transfer performance as a function of how likely the target task is according to the prior.}
  \label{fig:dipfdbsgraphplots}
\end{figure*}

To prove that our guesses are correct, we generate a set of $50$ source tasks for the three-room environment of Figure~\ref{f:rooms} by sampling both door locations uniformly in the allowed space, and solve all of them by directly minimizing the TD error as presented in Section~\ref{sec:td}. Then, we use our algorithms to transfer from $10$ source tasks sampled from the previously generated set. The average return over the last $50$ learning episodes as a function of the number of iterations is shown in Figure~\ref{f:two_doors}. Each curve is the result of $20$ independent runs, each one resampling the target and source tasks, with $95\%$ confidence intervals. Further details on the parameters adopted in this experiment are given in App. \ref{}. As expected, the no-transfer (NT) algorithm fails at learning the task in so few iterations due to the limited exploration provided by an $\epsilon$-greedy policy. On the other hand, all our algorithms achieve a significant speed-up and converge to the optimal performance in few iterations, with GVT being slightly slower. Interestingly, we notice that there is no advantage in adopting more than $1$ component for the posterior in MGVT. This result is intuitive since, as soon as the algorithm figures out which is the target task, all the components move towards the same region.

To better understand the differences between GVT and MGVT, we now consider transferring from a slightly different distribution than the one from which target tasks are drawn. We generate $50$ source tasks again but this time with the bottom door fixed at the center and the other one moving. Then, we repeat the previous experiment, allowing both doors to move when sampling target tasks. The results are shown in Figure~\ref{f:one_door}. Interestingly, MGVT seems almost unaffected by this change, proving that it has sufficient representation power to generalize to slightly different task distributions. The same does not hold for GVT, which now is not able to solve many of the sampled target tasks, as can be noticed from the higher variance. This result proves again that assuming Gaussian distributions can pose severe limitations in our transfer settings.

Finally, we analyze the transfer performance as a function of how likely the target task is according to the prior. We consider a two-room version of the environment of Figure~\ref{f:rooms}. Unlike before, we generate tasks by sampling the door position from a Gaussian with mean $5$ and standard deviation $1.8$, so that tasks with the door near the sides are very unlikely. Figure~\ref{f:likelihood} shows the performance reached by GVT and 1-MGVT at fixed iterations as a function of how likely the target task is according to such distribution. As expected, GVT achieves poor performance on very unlikely tasks, even after many iterations. In fact, estimating a single Gaussian distribution definitely entails some information loss, especially about the unlikely tasks. On the other hand, MGVT keeps such information and, consequently, performs much better. Perhaps not surprisingly, MGVT reaches the optimal performance in $4k$ iterations no matter what task is being solved.

@@

We now consider two well-known classic control environments: Cartpole and Mountain Car \citep{sutton1998reinforcement}. For both, we generate $20$ source tasks by uniformly sampling their physical parameters (cart mass, pole mass, pole length for Cartpole and car speed for Mountain Car). We parameterize $Q$-functions using neural networks with $1$-layer of $32$ hidden units for Cartpole and $64$ for Mountain Car. A better description of these two environments and their parameters is given in App. \ref{}. In this experiment, we use a Double Deep Q-Network \cite{van2016deep} to provide a stronger no-transfer baseline for comparison. The results (same settings of Section~\ref{sec:gw}) are shown in Figures~\ref{f:cartpole} and~\ref{f:mountaincar}. For Cartpole (Figure~\ref{f:cartpole}), all transfer algorithms are almost zero-shot. This result is expected since, although we vary the system parameters in a wide range, the optimal $Q$-values of states near the balanced position are almost the same for all tasks. On the contrary, in Mountain Car (Figure~\ref{f:mountaincar}) the optimal $Q$-functions become very different when changing the car speed. This phenomenon hinders the learning of GVT in the target task, while 1-MGVT achieves a good jump-start and converges in fewer iterations.

@@

\begin{figure*}[t]
  \begin{subfigure}[b]{0.15\textwidth}
    \includegraphics[trim=0.0cm 0.0cm 0.0cm 0.0cm,clip=true,height=2.6cm, width=2.0cm]{images/maze/maze6.png}
    \caption{}\label{f:maze}
  \end{subfigure}
  \begin{subfigure}[b]{0.27\textwidth}
    \includegraphics[trim=0.5cm 0cm 1.8cm 1.3cm,clip=true,height=2.9cm]{images/cartpole/lrev.pdf}
    \caption{}\label{f:cartpole}
  \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
    \includegraphics[trim=0.5cm 0cm 1.6cm 1.3cm,clip=true,height=2.9cm]{images/mountaincar/lrev.pdf}
    \caption{}\label{f:mountaincar}
  \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
    \includegraphics[trim=0.5cm 0cm 1.6cm 1.3cm,clip=true,height=2.9cm]{images/maze/lrev.pdf}
    \caption{}\label{f:maze_results}
  \end{subfigure}
  \caption{(a) Sample Maze (Maze 6), (b) transfer from $10$ sources in Cartpole, (c) transfer from $10$ source tasks in Mountain Car, and (d) transfer from $10$ sources in Mazes 6 and 13---see App. \ref{}.}
  \label{fig:ccmazeplots}
\end{figure*}
